---
title: "Simple linear regression: categorical predictor"
subtitle: "Notes and in-class exercises"
sidebar: false
---

# Notes

- You can download a template file to work with [here](../activity_templates/07_slr_cat_predictor.qmd).
- **File organization:** Save this file in the "Activities" subfolder of your "STAT155" folder.

## Learning goals

By the end of this lesson, you should be able to:

- Write a **model formula** for a simple linear regression model with a categorical predictor using **indicator variables**
- Interpret the **intercept** and **slope** coefficients in a simple linear regression model with a categorical predictor

## Readings and videos

Before class you should have read and/or watched:

- Reading: Section 3.9 in the [STAT 155 Notes](https://mac-stat.github.io/Stat155Notes/) only up through section 3.9.1 Indicator Variables
- Videos:
    - [Simple linear regression: categorical predictor](https://youtu.be/QPpp6Kpk2To) ([slides](https://drive.google.com/file/d/1fbSJW4DhVWxR8rykowPUoAZ6WfzuZ857/view?usp=sharing))
    - [R Code for Categorical Predictors](https://voicethread.com/share/14980390/)

## Review

So far we've focused on modeling a quantitative response/outcome as a function of a single quantitative predictor.

$$
E[\text{price} \mid \text{weight\_grams}] = \beta_0 + \beta_1 \text{weight\_grams}
$$

But when building models to explain why the outcome varies, we will likely have categorical predictors too. For example, the **quality** of the product might be encoded as "low" or "high".

Using a model like

$$
E[\text{price} \mid \text{quality}] = \beta_0 + \beta_1 \text{quality}
$$

doesn't make much sense because what does $\beta_1\,\text{quality}$ mean? How can we multiply a number by a word/category label? We can't!


<br><br><br><br>


For this reason, we make use of **indicator variables**. An indicator variable is a binary variable that only takes the values 0 or 1.

- Example: turn the `quality` variable into a `qualityHigh` indicator variable:
    - `qualityHigh = 1` if the quality of the product is "high"
    - `qualityHigh = 0` if the quality of the product is "low"

**Question:** Alternatively, we could have turned `quality` variable into a `qualityLow` indicator variable. What would the `qualityHigh` and `qualityLow` indicator variables look like for the following 4 cases?

price     quality     qualityHigh     qualityLow
------    --------    ------------    -----------
10          low
20          high
11          low
21          high


<br><br><br><br>


The model statement in terms of indicator variables would be:

$$
E[\text{price} \mid \text{quality}] = \beta_0 + \beta_1 \text{qualityHigh}
$$

- Note: for brevity, we can write "quality" instead of "qualityHigh" in the expected value (E[...] part) because we're using this to indicate that price depends on the quality variable as a whole rather than the specific indicator variables chosen.
    - If we listed all of the indicator variables inside the expected value, this would get long when a categorical variable has 3 or more categories

**Question:** In terms of the $\beta$ coefficients, what is the average price of a low quality product? What about a high quality product?




<br><br><br><br>


For categorical variables with more than 2 categories, we need to create multiple indicator variables. To do this we pick one category to be the **reference category** and create indicator variables for the other categories.

For example, if quality had "low", "medium", and "high" categories, we could pick "low" to be the reference category (conventional for the "first" category to be the reference) and create `qualityMedium` and `qualityHigh` indicator variables:

$$
E[\text{price} \mid \text{quality}] = \beta_0 + \beta_1 \text{qualityMedium} + \beta_2 \text{qualityHigh}
$$

**Question:** In terms of the $\beta$ coefficients, what is the average price of a low quality product? A medium quality product? A high quality product?





<br><br><br><br>


# Exercises

**Context:** Today we'll explore data on thousands of diamonds to understand how physical characteristics relate to price. Read in the data below.

```{r warning=FALSE, message=FALSE}
# Load packages and import data
library(ggplot2)
library(dplyr)

data(diamonds)

# A little bit of data wrangling code - let's not focus on this for now
diamonds <- diamonds %>% 
    mutate(
        cut = factor(cut, ordered = FALSE),
        color = factor(color, ordered = FALSE),
        clarity = factor(clarity, ordered = FALSE)
    )
```

For the first several exercises, our focus will be on the relationship between diamond price and cut.

## Exercise 1: Get to know the data

Write R code to answer the following:

- How many cases and variables do we have? What does a case represent?
- What do the first few rows of the data look like?
- Construct and interpret two different visualizations of the `price` variable.
- Construct and interpret a visualization of the `cut` variable.



## Exercise 2: Visualizations

Start by visualizing this relationship of interest, that between `price` and `cut`.

a. The appropriate plot depends upon the type of variables we're plotting. When exploring the relationship between a quantitative response (ridership) and a quantitative predictor (temperature), a scatterplot was an effective choice. After running the code below, explain why a scatterplot is not effective for exploring the relationship between ridership and our *categorical* `cut` predictor.

```{r eval=FALSE}
# Try a scatterplot
ggplot(diamonds, aes(y = price, x = cut)) + 
    geom_point()
```

b. Separately run each chunk below, with two plots. Comment (`#`) on what changes in the code / output.

```{r eval=FALSE}
# Univariate boxplot
ggplot(diamonds, aes(y = price)) + 
    geom_boxplot()
```

```{r eval=FALSE}
# ???
ggplot(diamonds, aes(y = price, x = cut)) + 
    geom_boxplot()
```

```{r eval=FALSE}
# Univariate density plot
ggplot(diamonds, aes(x = price)) + 
    geom_density()
```

```{r eval=FALSE}
# ???
ggplot(diamonds, aes(x = price, color = cut)) + 
    geom_density()
```

```{r eval=FALSE}
# Univariate histogram
ggplot(diamonds, aes(x = price)) + 
    geom_histogram()
```

```{r eval=FALSE}
# ???
ggplot(diamonds, aes(x = price)) + 
    geom_histogram() + 
    facet_wrap(~ cut)
```

c. Do you notice anything interesting about the relationship between price and cut? What do you think might be happening here?



## Exercise 3: Numerical summaries

Let's follow up our plots with some numerical summaries.

a. To warm up, first calculate the mean `price` across all diamonds.

```{r eval=FALSE}
diamonds %>% 
    ___(mean(___))
```

b. To summarize the trends we observed in the grouped plots above, we can calculate the mean `price` for each type of `cut`. This requires the inclusion of the `group_by()` function:

```{r eval=FALSE}
# Calculate mean price by cut
diamonds %>% 
    group_by(cut) %>% 
    ___(mean(___))
```

c. Examine the group mean measurements, and make sure that you can match these numbers up with what you see in the plots.

d. Based on the results above, we can see that, on average, diamonds with a "Fair" cut tend to cost more than higher-quality cuts. Let's construct a new variable named `cutFair`, using on the following criteria:

- `cutFair` = 1 if the diamond is of Fair cut
- `cutFair` = 0 otherwise (any other value of cut (Good, Very Good, Premium, Ideal))

```{r eval=FALSE}
diamonds <- diamonds %>%
  mutate(cutFair=ifelse(___ == "Fair", __, __))
```

Variables like `cutFair` that are coded as 0/1 to numerically indicate if a categorical variable is at a particular state are known as an **indicator variable**. You will sometimes see these referred to as a "binary variable" or "dichotomous variable"; you may also encounter the term "dummy variable" in older statistical literature.

e. Now, let's calculate the group means based on the new `cutFair` indicator variable:

```{r eval=FALSE}
diamonds %>% 
    group_by(cutFair) %>% 
    summarize(mean(price))
```

## Exercise 4: Modeling trend using a categorical predictor with exactly 2 categories 

Next, let's *model* the trend in the relationship between the `cutFair` and `price` variables using a simple linear regression model:

```{r eval=FALSE}
# Construct the model
diamond_mod0 <- lm(price ~ cutFair, data = diamonds)

# Summarize the model
coef(summary(diamond_mod0))
```


Compare these results to the output of exercise 3e. What do you notice? How do you interpret the intercept and `cutFair` coefficient terms from this model?

> your answer here

## Exercise 5: Modeling trend using a categorical predictor with >2 categories

Using a single binary predictor like the `cutFair` indicator variable is useful when there are two clearly delineated categories. However, the `cut` variable actually contains 5 categories! Because we've collapsed all non-Fair classifications into a single category (i.e. `cutFair = 0`), the model above can't tell us anything about the difference in expected price between, say, *Premium* and *Ideal* cuts. The good news is that it is very straightforward to model categorical predictors with >2 categories. We can do this by using the `cut` variable as our predictor: 

```{r eval=FALSE}
# Construct the model
diamond_mod <- lm(price ~ cut, data = diamonds)

# Summarize the model
coef(summary(diamond_mod))
```

a. Even though we specified a single predictor variable in the model, we are seeing 4 coefficient estimates--why do you think this is the case?

> your answer here

---

NOTE: We see 4 indicator variables (for Good, Very Good, Premium, and Ideal), but we do not see `cutFair` in the model output. This is because `Fair` is the **reference level** of the `cut` variable (it's first alphabetically). You'll see below that it is, indeed, still in the model. You'll also see why the term "reference level" makes sense!
  
---

b. After examining the summary table output from the code chunk above, complete the model formula:

\

E[price | cut] = ___ +/- ___ cutGood +/- ___ cutVery Good +/- ___ cutPremium +/- ___ cutIdeal

\   


## Exercise 6: Making sense of the model

Recall our model:  E[price | cut] = 4358.7578 - 429.8933 cutGood - 376.9979 cutVery Good + 225.4999 cutPremium - 901.2158 cutIdeal

a. Use the model formula to calculate the expected/typical price for diamonds of *Good* cut.

b. Similarly, calculate the expected/typical price for diamonds of *Fair* cut.

c. Re-examine these 2 calculations. Where have you seen these numbers before?!


## Exercise 7: Interpreting coefficients

Recall that our model formula is *not a formula for a line*. Thus we can't interpret the coefficients as "slopes" as we have before. Taking this into account and reflecting upon your calculations above...   

a. Interpret the intercept coefficient (`4358.7578`) in terms of the data context. *Make sure to use non-causal language, include units, and talk about averages rather than individual cases.*

b. Interpret the `cutGood` and `cutVery Good` coefficients (`-429.8933` and `-376.9979`) in terms of the data context. Hint: where did you use these value in the prediction calculations above?



## Exercise 8: Modeling choices (CHALLENGE)

Why do we fit this model in this way (using 4 indicator variables `cutGood`,  `cutVery Good`, `cutPremium`, `cutIdeal`)? Instead, suppose that we created a single variable `cutCat` that gave each category a numerical value: 0 for Fair, 1 for Good, 2 for Very Good, 3 for Premium, and 4 for Ideal.

How would this change things? What are the pros and cons of each approach?



## Reflection

Through the exercises above, you learned how to build and interpret models that incorporate a categorical predictor variable. For the benefit of your future self, summarize how one can interpret the coefficients for a categorical predictor.

> **Response:** Put your response here.



## Render your work

- Click the "Render" button in the menu bar for this pane (blue arrow pointing right). This will create an HTML file containing all of the directions, code, and responses from this activity. A preview of the HTML will appear in the browser.
- Scroll through and inspect the document to check that your work translated to the HTML format correctly.
- Close the browser tab.
- Go to the "Background Jobs" pane in RStudio and click the Stop button to end the rendering process.
- Navigate to your "Activities" subfolder within your "STAT155" folder and locate the HTML file. You can open it again in your browser to double check.



# Additional Practice

## Exercise 9: Diamond color

Consider modeling `price` by `color`.

- Before creating a visualization that shows the relationship between price and color, write down what you expect the plot to look like. Then construct and interpret an apporpriate plot.
- Compute the average price for each color.
- Fit an appropriate linear model with `lm()` and display a short summary of the model.
- Write out the model formula from the above summary.
- Which color is the reference level? How can you tell from the model summary?
- Interpret the intercept and two other coefficients from the model in terms of the data context.



## Exercise 10: Diamond clarity

If you want more practice, repeat the steps from Exercise 8 for the `clarity` variable.



<br><br><br><br><br><br>



# Solutions

## Exercise 1: Get to know the data

- A case represents a single diamond.
- The distribution of price is right skewed with considerable high outliers. The right skew is evidenced by the mean price (\$3932) being much higher than the median price (\$2401).
- Most diamonds in this data are of Good cut or better. Ideal cut diamonds are the most common with each succesive grade being the next most common.


```{r}
dim(diamonds)

head(diamonds)

# Visualize price (outcome variable)
ggplot(diamonds, aes(x = price)) +
    geom_histogram()
ggplot(diamonds, aes(y = price)) +
    geom_boxplot()
diamonds %>%
    summarize(mean(price), median(price), sd(price))

# Visualize cut (predictor variable)
ggplot(diamonds, aes(x = cut)) +
    geom_bar()
diamonds %>% 
    count(cut)
```




## Exercise 2: Visualizations

Start by visualizing this relationship of interest, that between `price` and `cut`.

a. We just don't see anything clearly on a scatterplot. With the small number of unique values of the predictor variable, all of the points are bunched up on each other.

```{r}
# Try a scatterplot
ggplot(diamonds, aes(y = price, x = cut)) + 
    geom_point()
```

b. Separately run each chunk below, with two plots. Comment (`#`) on what changes in the code / output.

```{r}
# Univariate boxplot
ggplot(diamonds, aes(y = price)) + 
    geom_boxplot()
```

```{r}
# Separate boxes by category
ggplot(diamonds, aes(y = price, x = cut)) + 
    geom_boxplot()
```

```{r}
# Univariate density plot
ggplot(diamonds, aes(x = price)) + 
    geom_density()
```

```{r}
# Separate density plots by category
ggplot(diamonds, aes(x = price, color = cut)) + 
    geom_density()
```

```{r}
# Univariate histogram
ggplot(diamonds, aes(x = price)) + 
    geom_histogram()
```

```{r}
# Separate histograms by category
ggplot(diamonds, aes(x = price)) + 
    geom_histogram() + 
    facet_wrap(~ cut)
```

c. The relationship between price and cut seems to be opposite what we would expect. The diamonds with the best cut (Ideal) have the lowest average price, and the ones with the worst cut (Fair) are woth the most. Maybe something else is different between the diamonds with the best and worst cuts...size maybe?



## Exercise 3: Numerical summaries

Let's follow up our plots with some numerical summaries.

a. Mean `price` across all diamonds:

```{r}
diamonds %>% 
    summarize(mean(price))
```

b. Mean `price` for each type of `cut`:

```{r}
diamonds %>% 
    group_by(cut) %>% 
    summarize(mean(price))
```

c. Group means should reflect what you see in the plots (easiest to see in the boxplots)

d. Create our new `cutFair` variable:

```{r}
diamonds <- diamonds %>%
  mutate(cutFair=ifelse(cut == "Fair", 1, 0))
```

e. Calculate the group means based on this new variable

```{r}
diamonds %>% 
    group_by(cutFair) %>% 
    summarize(mean(price))
```

## Exercise 4: Modeling trend using a categorical predictor with exactly 2 categories 

```{r}
# Construct the model
diamond_mod0 <- lm(price ~ cutFair, data = diamonds)

# Summarize the model
coef(summary(diamond_mod0))
```

The intercept is the expected value (mean) of the price for all diamonds with a cut quality that isn't Fair (Good, Very Good, Premium, or Ideal, i.e. when `cutFair` = 0)--the same as we saw in exercise 3e. 

c. When we add the intercept and coefficient for `cutFair`, we get `3919.69 + 439.06 = 4358.75`--this is the mean price for all diamonds with a Fair cut quality that we saw in exercise 3e! Therefore, the coefficient of `cutFair` (439.06) is interpreted as the difference between the mean value of diamonds with a Fair cut quality and the mean value of diamonds with a higher cut quality.

## Exercise 5: Modeling trend using a categorical predictor with >2 categories 

```{r}
# Construct the model
diamond_mod <- lm(price ~ cut, data = diamonds)

# Summarize the model
coef(summary(diamond_mod))
```

a. We are seeing 4 coefficient estimates because each category is being assigned to a separate indicator variable--`cutGood` = 1 when `cut == "Good"` and 0 otherwise, `cutVery Good` = 1 when `cut == "Very Good" and 0 otherwise, and so on.

b. E[price | cut] = 4358.7578 - 429.8933 cutGood - 376.9979 cutVery Good + 225.4999 cutPremium - 901.2158 cutIdeal



## Exercise 6: Making sense of the model

a. Expected/typical price for diamonds of *Good* cut:

E[price | cut] = 4358.7578 - 429.8933 * 1 - 376.9979 * 0 + 225.4999 * 0 - 901.2158 * 0 = 4358.7578 - 429.8933 = \$3928.865

```{r}
predict(diamond_mod, newdata = data.frame(cut = "Good"))
```

b. Expected/typical price for diamonds of *Fair* cut:

E[price | cut] = 4358.7578 - 429.8933 * 0 - 376.9979 * 0 + 225.4999 * 0 - 901.2158 * 0 = \$4358.7578

```{r}
predict(diamond_mod, newdata = data.frame(cut = "Fair"))
```

c. These come from our group mean calculations in Exercise 3b! The predicted value for diamonds of *Fair* cut is also the same as what we obtained using the SLR model in exercise 4 with only a single `cutFair` indicator variable.


## Exercise 7: Interpreting coefficients

Recall that our model formula is *not a formula for a line*. Thus we can't interpret the coefficients as "slopes" as we have before. Taking this into account and reflecting upon your calculations above...   

a. The average price of a Fair cut diamonds is \$4358.7578.

b.
    - Interpretation of `cutGood` coefficient: On average, Good cut diamonds are worth \$429.89 less than Fair cut diamonds.
    - Interpretation of `cutVery Good` coefficient: On average, Very Good cut diamonds are worth \$377.00 less than Fair cut diamonds.



## Exercise 8: Modeling choices (CHALLENGE)

Why do we fit this model in this way (using 4 indicator variables `cutGood`,  `cutVery Good`, `cutPremium`, `cutIdeal`)? Instead, suppose that we created a single variable `cutCat` that gave each category a numerical value: 0 for Fair, 1 for Good, 2 for Very Good, 3 for Premium, and 4 for Ideal.

- If we used 0-4 instead of creating indicator variables, we would be constraining the change from 0 to 1, from 1 to 2, etc. to always be of the same magnitude. That is, a 1 unit change in the cut variable would always have the same change in price in our model.
- Using separate indicator variables allows the difference between subsequent categories to be different, which allows our model to be a bit more nuanced. It is possible to take nuance too far though. For example, in our previous investigations of bikeshare data, we modeled ridership versus temperature. We treated temperature as a quantitative predictor. Imagine if we had created an indicator variable for each unique temperature in the data---that would be so many variables! Having so many variables creates a very complex model which can be hard to make sense of. (These ideas are addressed further in STAT 253: Statistical Machine Learning!)



## Exercise 9: Diamond color

Consider modeling `price` by `color`.

- The best color diamonds are J, and worst are D. We would expect D diamonds to have the lowest price and increase steadily as we get to J. This is in fact what we see in the boxplots.

```{r}
ggplot(diamonds, aes(x = color, y = price)) +
    geom_boxplot()

diamonds %>% 
    group_by(color) %>% 
    summarize(mean(price))
```


- We fit a linear model and obtain the model formula: E[price | color] = 3169.95 - 93.20 colorE + 554.93 colorF + 829.18 colorG + 1316.72 colorH + 1921.92 colorI + 2153.86 colorJ

```{r}
diamond_mod2 <- lm(price ~ color, data = diamonds)

coef(summary(diamond_mod2))
```


- Color D is the reference level because we don't see its indicator variable in the model output.
- Interpretation of the intercept: Diamonds with D color cost \$3169.95 on average.
- Interpretation of the `colorE` coefficient: Diamonds with E color cost \$93.20 **less** than D color diamonds on average.
- Interpretation of the `colorF` coefficient: Diamonds with F color cost \$554.93 **more** than D color diamonds on average.



## Exercise 10: Diamond clarity

We see the unexpected result that diamonds of better clarity (VS1 and higher) have lower average prices. In fact the best clarity diamonds (VVS1 and IF) have the *lowest* average prices. What might be going on? What if the most clear diamonds were also quite small...

```{r}
ggplot(diamonds, aes(x = clarity, y = price)) +
    geom_boxplot()

diamonds %>% 
    group_by(clarity) %>% 
    summarize(mean(price))

diamond_mod3 <- lm(price ~ clarity, data = diamonds)

coef(summary(diamond_mod3))
```



